Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-08 Manuscript/manuscript.Rmd
Review ???

To investigate whether we could alleviate resolution limitations of conventional TAD and loop boundary calling algorithms, we applied our RF model to predict boundaries from base pair (bp) resolution data. [??? Write how your vector Y and annotation matrices change;!!!added]. At this point, we opted to implement an optimally built whole genome RF model. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements). Here, $Y$ was of length $N$ where $N=n_1+n_2+\cdots+n_{22}$, the total number of bins across each chromosome. The data matrix $\textbf{X}$ had dimensions $N \times p_s$, where $p_s \subset P$, a subset of the most predictive genomic elements. Given that each chromosome is composed of tens, in some cases hundreds, of millions of base pairs, we used the same whole genome RF model to predict boundaries from chromosome-specific bp resolution data.

@@

First, we extracted the vector $p_{n_{i}}$ that denoted the probability of being a boundary point for each chromosome-specific base pair from our RF model, where $n_{i}$ was the length of chromosome $i$. We then applied *DBSCAN* (version 1.1-5) to the matrix of pairwise distances between base pairs with $p_{n_{i}}=1$. The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts* [@hahsler2019dbscan]. The eps parameter was assigned as the respective data resolution. The heuristic being that base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3. The resulting clusters of density-reachable base pairs identified by DBSCAN were termed *predicted TAD boundary regions* (PTBR) [??? This deserves a schematic figure, before Figure 5. And part of it may be in Results;!!!added PTBR to example of Figure 5 A] [??? I would switch the notation. First, you detect REGIONS. They can be adjacent, clustered together, forming AREA. Within this area, you can do pam clustering, and identify POINT;!!!editted, used term "region" instead of "area"]. To precisely identify the most representative base pair among each PTBR we performed an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. Thusly, the corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base pair within each clustered PTBR.

>>>>

To investigate whether we could alleviate resolution limitations of conventional TAD and loop boundary calling algorithms, we applied our RF model to predict boundaries from base pair (bp) resolution data. [??? Write how your vector Y and annotation matrices change;!!!added ???You describe whole-genome model. It should have been described early on. Here, describe chromosome-specific bp-resolution model]. At this point, we opted to implement an optimally built whole genome RF model. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements). Here, $Y$ was of length $N$ where $N=n_1+n_2+\cdots+n_{22}$, the total number of bins across each chromosome. The data matrix $\textbf{X}$ had dimensions $N \times p_s$, where $p_s \subset P$, a subset of the most predictive genomic elements. Given that each chromosome is composed of tens, in some cases hundreds, of millions of base pairs, we used the same whole genome RF model to predict boundaries from chromosome-specific bp resolution data.

@@

First, we extracted the vector $p_{n_{i}}$ that denoted the probability of being a boundary point for each chromosome-specific base pair from our RF model, where $n_{i}$ was the length of chromosome $i$. We then applied *DBSCAN* (version 1.1-5) to the matrix of pairwise distances [??? Is that true? Matrix is 2-dimensional] between base pairs with $p_{n_{i}}=1$. The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts* [@hahsler2019dbscan]. The eps parameter was assigned as the respective data resolution [??? Meaning?]. The heuristic being that base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3 [??? Meaning?]. The resulting clusters of density-reachable base pairs identified by DBSCAN were termed *predicted TAD boundary regions* (PTBR) [??? Use simple words: "boundary region" (many bases clustered by DBSCAN), then "boundary base" (selected by PAM)]. To precisely identify the most representative base pair among each PTBR we performed an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. Thusly, the corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base pair within each clustered PTBR.
